Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-04-18 closed_loop_planning.tex
[MGa] updated closed loop and mod open loop
Restructured eqs in \subequations, ...

The approach is based on the three following steps: i) A nominal time-optimal trajectory is planned: this will be referred to as nominal feed-forward optimal trajectory. This is the trajectory a driver should follow in the ideal situation of a world without uncertainty. Therefore, in this step the mean trajectory is the actual trajectory a deterministic vehicle dynamics would trace. ii) A closed-loop controller mimicking the driver's action is computed which stabilizes the nominal feed-forward optimal trajectory computed at step i). We assume a discrete time-varying Linear Quadratic Regulator (LQR) controller is a good approximation of how an expert driver can track a prescribed trajectory. iii) In the last step the stochastic framework is reintroduced and the time-optimal planning problem is robustified using the Lyapunov framework. The main idea is to \emph{re-plan} a mean trajectory of a stochastic system such that, a time-varying LQR controller that aims at stabilizing it, is able to properly tame the propagation of the uncertainty. This amounts to ensuring that the robustified path constraints are satisfied. More specifically, this implies two key aspects: first, that the covariance matrix is propagated accounting explicitly for the controlled system dynamics; and second, that the path constraints are robustly satisfied by incorporating the covariance estimation propagated through the closed-loop controlled system. Some aspects that may initially appear technical are in fact essential. In this third step, not only are states and controls re-planned, but also the time-varying LQR gains. In fact, they need to explicitly account for the changed requirements associated with stabilizing a trajectory different from the nominal one, which must now satisfy the robustified constraints.
\subsection{Step 1 -- Nominal feed-forward time-optimal trajectory planning}
\label{sec:nominalFF}

@@

Given the nominal trajectory $(\hbmu_k, \hbu_k)$ over $k=0,\ldots, N$, and introduced the deviation variables $\bbmu_k=\bmu_k - \hbmu_k$ and $\bbu = \bu_k - \hbu_k$, we design a linear state-feedback law $\bbu_k= -\hbK_k \bbmu_k$, or equivalently $\bu_k = \hbu_k - \hbK_k (\bmu_k - \hbmu_k)$, to stabilize the deviations along the nominal trajectory. To compute the state-feedback matrices $\hbK_k$, we linearize the system around $(\hbmu_k, \hbu_k)$, such that $\bbmu_{k+1} \approx \hbA_{k}\bbmu_k + \hbB_k \bbu_k$, where $\hbA_k = f_{,\bmu}(\hbmu_k,\hbu_k)$ and $\hbB_k = f_{,\bu}(\hbmu_k,\hbu_k)$, and define the quadratic regulator (tracking) cost function
\begin{align}
J = \bbmu_N^T \bW_N \bbmu_N + \sum_{k=0}^{N-1} \big( \bbmu_k^{T} \bW_k \bbmu_k + \bbu_k^T \bR_k \bbu_k \big),
\end{align}
where $\bW_N=\bW_N^T \succ 0$, $\bW_k=\bW_k^T \succeq 0$, $\bR_k=\bR_k^T \succ 0$. It can be shown that the optimal cost-to-go is given by $J_k^*(\bbmu_k) = \bbmu_k^T \bS_k \bbmu_k$, with $\bS_k = \bS_k^T \succ 0$, where $\bS_k$ can be computed through the (backward) Riccati recursion for $k=N-1,\ldots, 0$
\begin{align}
\bS_N &= \bW_N,\\ 
\bS_{k} &= \bW_k + \hbA_k^T \bS_{k+1} \hbA_k - \hbA_k^T \bS_{k+1}\hbB_k \hbK_k,\quad k={N-1, \ldots, 0}
\end{align}
where  $\hbK_k =\big( \bR_k+\hbB_k^T \bS_{k+1} \hbB_k \big)^{-1} \hbB_k^T \bS_{k+1} \hbA_k$. The optimal (stabilizing) feedback policy is given by $\bbu_k=-\hbK_k \bbmu_k$. We assume that the gain sequence $\{\hbK_k\}$, $k=0,\ldots, N$ represents the control strategy that an expert driver would employ to optimally stabilize the system along the nominal trajectory.

\subsection{Step 3 -- Closed-loop robustified planning}
\label{sec:CLrobustified}

%\begin{align}
%    \underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}} \quad & J_k(\bmu_k,\bxi_k, \bu_k) \label{eq:ROCPcost}\\
%       \text{s.t.}\notag\\
%          & f_0(x) - t \le 0 \quad \phantom{i=1, \dots, m}\\
%          & f_i(x) \le 0, \quad i=1, \dots, m \quad i=1, \dots, m\\
%          & h_i(x) = 0,   \quad i=1, \dots, p,
%\end{align}

@@

where~\eqref{eq:ROCPdP} denotes the discretized form for time evolution of the covariance matrix under a stabilizing controller of the form $\bbu_k=-\bK_k \bbmu_k$. As apposed to~\eqref{eq:OCPdP}, $\tbP$ would evolve, in continuous time, according to $\dot{\tbP}(t) = (\bA(t)-\bB(t) \bK(t)) \tbP(t) + \tbP(t) (\bA(t)-\bB(t) \bK(t))^T + \bQ(t)$. To be continued...

>>>>

The approach is based on the three following steps: i) A nominal time-optimal trajectory is planned: this will be referred to as nominal feed-forward optimal trajectory. This is the trajectory a driver should follow in the ideal situation of a world without uncertainty. In this step the mean trajectory is the actual trajectory a deterministic vehicle dynamics would trace.

ii) A closed-loop controller mimicking the driver's action is computed which stabilizes the nominal feed-forward optimal trajectory computed at step i). We assume a discrete time-varying Linear Quadratic Regulator (LQR) controller is a good approximation of how an expert driver can track a prescribed trajectory.

iii) In the last step the stochastic framework is reintroduced and the time-optimal planning problem is robustified using the Lyapunov framework. The main idea is to \emph{re-plan} a mean trajectory of a stochastic system such that, a time-varying LQR controller that aims at stabilizing it, is able to properly tame the propagation of the uncertainty. This amounts to ensuring that the robustified path constraints are satisfied. More specifically, this accounts for two key aspects: firstly, that the covariance matrix is propagated accounting explicitly for the controlled system dynamics; and secondly, that the path constraints are robustly satisfied by incorporating the covariance estimation propagated through the closed-loop controlled system. Some aspects that may appear a bit technical are in fact essential. In this third step, not only are states and controls re-planned, but also the time-varying LQR gains. As a matter of fact, they need to explicitly account for the changed requirements associated with stabilizing a trajectory different from the nominal one, which must now satisfy the robustified constraints.
\subsection{Step 1 -- Nominal feed-forward time-optimal trajectory planning}
\label{sec:nominalFF}
The first step consists in planning a nominal feed-forward trajectory. This assumes deterministic system dynamics, with \( (\hbmu_k, \hbu_k) \) denoting the differential states and controls, and \( \hbz_k \) the algebraic states, for \( k = 0, \ldots, N \). To avoid redundancy, we note that the NLP to be solved is a special case of~\eqref{eq:DOCP}, where~\eqref{eq:DOCPdP},~\eqref{eq:DOCPdPIC}, and the backoff terms are omitted, since the goal is to compute a nominal trajectory.

@@

Given the nominal trajectory $(\hbmu_k, \hbu_k)$ over $k=0,\ldots, N$, and introduced the deviation variables $\bbmu_k=\bmu_k - \hbmu_k$ and $\bbu = \bu_k - \hbu_k$, we design a linear state-feedback law $\bbu_k= -\hbK_k \bbmu_k$, or equivalently $\bu_k = \hbu_k - \hbK_k (\bmu_k - \hbmu_k)$, to mitigate disturbances while tracking the nominal trajectory. To compute the state-feedback matrices $\hbK_k$, we linearize the system around $(\hbmu_k, \hbu_k)$, such that $\bbmu_{k+1} \approx \hbA_{k}\bbmu_k + \hbB_k \bbu_k$, where $\hbA_k = f_{,\bmu}(\hbmu_k,\hbu_k)$ and $\hbB_k = f_{,\bu}(\hbmu_k,\hbu_k)$, and define the quadratic regulator (tracking) cost function
\begin{align}
J = \bbmu_N^T \bW_N \bbmu_N + \sum_{k=0}^{N-1} \big( \bbmu_k^{T} \bW_k \bbmu_k + \bbu_k^T \bR_k \bbu_k \big),
\end{align}
where $\bW_N=\bW_N^T \succ 0$, $\bW_k=\bW_k^T \succeq 0$, $\bR_k=\bR_k^T \succ 0$. It can be shown that the optimal cost-to-go is given by $J_k^*(\bbmu_k) = \bbmu_k^T \bS_k \bbmu_k$, with $\bS_k = \bS_k^T \succ 0$, where $\bS_k$ can be computed through the (backward) Riccati recursion for $k=N-1,\ldots, 0$
\begin{align}
\bS_N &= \bW_N,\\ 
\bS_{k} &= \bW_k + \hbA_k^T \bS_{k+1} \hbA_k - \hbA_k^T \bS_{k+1}\hbB_k \hbK_k,\quad k={N-1, \ldots, 0}
\end{align}
where  $\hbK_k =\big( \bR_k+\hbB_k^T \bS_{k+1} \hbB_k \big)^{-1} \hbB_k^T \bS_{k+1} \hbA_k$. The optimal (stabilizing) feedback policy is given by $\bbu_k=-\hbK_k \bbmu_k$. We assume that the gain sequence $\{\hbK_k\}$, $k=0,\ldots, N$ represents the control strategy that an expert driver would employ to optimally stabilize the system along the nominal trajectory.

\subsection{Step 3 -- Closed-loop robustified planning}
\label{sec:CLrobustified}
%\begin{align}
%    \underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}} \quad & J_k(\bmu_k,\bxi_k, \bu_k) \label{eq:ROCPcost}\\
%       \text{s.t.}\notag\\
%          & f_0(x) - t \le 0 \quad \phantom{i=1, \dots, m}\\
%          & f_i(x) \le 0, \quad i=1, \dots, m \quad i=1, \dots, m\\
%          & h_i(x) = 0,   \quad i=1, \dots, p,
%\end{align}
As the third and last step, a closed-loop robustified planning problem is set up. Its aim is the definition of a mean trajectory of a stochastic system such that, working in tandem with state-feedback controller that aims at stabilizing it, it is able to mitigate the perturbations and reduce the propagation of uncertainty. This helps to guarantee that the robustified path constraints can be satisfied over the entire planning horizon. The discretized version of this problem takes the form of the following NLP:
\begin{subequations}\label{eq:ROCP}

@@

\end{subequations} 
Here, \eqref{eq:ROCPdP} denotes the discretized form for time evolution of the covariance matrix under a stabilizing controller of the form $\bbu_k=-\bK_k \bbmu_k$. To mark the difference with~\eqref{eq:OCPdP}, a new symbol $\tbP$ is introduced to the denote the covariance matrix in this case. In continuous time, $\tbP$ evolves according to $\dot{\tbP}(t) = (\bA(t)-\bB(t) \bK(t)) \tbP(t) + \tbP(t) (\bA(t)-\bB(t) \bK(t))^T + \bQ(t)$. To be continued...

### 2025-04-18 open_loop_planning.tex
[MGa] updated closed loop and mod open loop
Restructured eqs in \subequations, ...

\section{Open-loop planning via H-steps ahead predictions}
\label{sec:open_loop_planning}
The proposed approach requires the presence of $H+1$ versions of the matrix $\bP$ at each grid point of the discretized problem. One of these versions is initialized with $\bar{\bP}_0$, while the others represent the evolution of matrices that were initialized at previous grid points, from 1 to $H$ steps earlier. This ensures that at each step it is possible to find a covariance matrix evolved for $H$ steps and use it to robustify the constraints. The initialization of the covariance matrix is necessary due to its divergent dynamics. In fact, without a feedback control, the error on the position and orientation of the body reference frame can only increase along the track, compromising the conditioning of the matrix.

%\subsection{Discretization via direct collocation}
%\label{sec:openloopdiscretization}
Introducing the versioning of the covariance matrix approach requires a modification of the Equations described in Section~\ref{sec:discretization}. In particular Eqs.~\eqref{eq:DOCPdP}, \eqref{eq:DOCPdPIC}, and \eqref{eq:DOCPconstraints} are replaced by Eqs.~\eqref{eq:DOCPdPopenloop}, \eqref{eq:DOCPdPinitopenloop}, and \eqref{eq:DOCPconstraintsopenloop}. Eq.~\eqref{eq:DOCPdPopenloop} collects the continuity and collocation Equations for all the versions of the covariance matrix that have propagated from the previous step. To account for the evolution of $H$ matrices, it is necessary to introduce $H\times d$ matrices $\bSi_k^j=(\bSi_{k,1}^j, \ldots, \bSi_{k,d}^j)$, which represent the values of each matrix version at the collocation points.

The specific formulation is expressed as follows:
\begin{alignat}{3}
	\underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k}{\text{minimize}} \,
	& & & J_k(\bmu_k,\bxi_k, \bu_k) & & \label{eq:DOCPcost} \\
	\hspace*{-2.0 cm}\text{s.t.} \quad
	& \bzero      & = & \; \bPsimu_k(\bmu_{k-1},\bmu_k,\bxi_k, \bu_k,\bz_k),
	& \quad & k = 1,\ldots, N \label{eq:DOCPdyn} \\
	& \bmu_0      & = & \; \bar{\bmu}_0
	& & \label{eq:DOCPdynIC} \\
	& \bzero      & = & \; \bPsiP_k(\bmu_k,\bxi_k, \bu_k, \bP_{k-1}^{j-1},\bP_k^j,\bSi_k^j,\bz_k),
	& \quad & \substack{k = 1,\ldots, N; \; \\ j=1,\ldots,H;\;} \label{eq:DOCPdPopenloop} \\
	& \bP_k^0       & = & \; \bar{\bP}_0 \succeq 0
	& \quad & k = 1,\ldots, N; \; \label{eq:DOCPdPinitopenloop} \\
	& \bzero      & = & \; \bOm_k(\bmu_k,\bxi_k, \bu_k,\bz_k),
	& \quad & k = 0,\ldots, N \label{eq:DOCPpath} \\
	& 0           & \geq & \; h_i(\bmu_k, \bu_k, \bz_k) + \be_i(\bmu_k, \bu_k, \bP_k^H, \bz_k),
	& \quad & k = 1,\ldots, N;\; i \in \calI \label{eq:DOCPconstraintsopenloop}
\end{alignat}

Eq.~\eqref{eq:DOCPdPinitopenloop} is used to initialize the correct version at the $k$-th step while Eq.~\eqref{eq:DOCPconstraintsopenloop} specify that the back-off term of the constraints is evaluated with the most propagated covariance matrix version, $\bP^H_k$.
A schematic representation of the approach used to manage the $H+1$ versions of the covariance matrix is shown in Figure \ref{fig:DOCPgrid}, where the dashed rectangle represent the $k$-th step at which the constraints are being formulated. At each grid point, two versions of the covariance matrix are highlighted: $\bP^0_k$ (red node), which is the version to be initialized, and $\bP^H_k$ (green node), which is the most propagated version used to formulate the robust constraints in Eq.~\eqref{eq:DOCPconstraintsopenloop}.

@@

	\end{tikzpicture}
	\caption{Schematic representation of the continuity and initialization Equations for the $H+1$ versions of the covariance matrix introduced in the discretized OCP. The dashed rectangle indicate the $k$-th step, in which the constraints are formulated. Each node represents a covariance matrix, labeled such that the subscript denotes the current grid step, while the superscript indicates the number of propagation steps it has undergone. At each step, two nodes are highlighted: the covariance matrix to be initialized (red node) and the matrix that has been propagated for $H$ steps (green node), which is used to formulate the robust constraints.}
	\label{fig:DOCPgrid}
\end{figure}

\subsection{Robust adherence constraint}

>>>>

\section{Open-loop planning via H-steps ahead predictions}
\label{sec:open_loop_planning}
In this section, we address the problem of planning under uncertainty for a stochastic system, where the covariance matrix is propagated in a feed-forward fashion for a fixed prediction horizon of $H$ steps at each grid point $k$ of the discretized trajectory. In the planning problem, we conservatively evaluate the maximum propagation of the covariance, that is, the matrix evolved over $H$ steps, to ensure the tightest possible enforcement of constraints. This maximizes the robustness of the resulting solution, as the constraint backoff is evaluated in the worst-case scenario.

To enable this, we associate multiple versions of the covariance matrix with each grid point. Specifically, $H+1$ instances of the matrix $\bP$ are maintained at each step. To this sake we introduce the notation $\bP_k^j$, where $j = 0, \ldots, H$, and $\bP_k^j$ represents the version of the covariance matrix at step $k$ that was initialized $j$ steps earlier. Hence, $\bP_k^0$ originates at node $k$ itself and will be propagated forward and employed in subsequent steps (exactly when evolved as $\bP_{k+H}^H$ ), while $\bP_k^H$ is the instance that originated $H$ steps earlier and is the most propagated one at step $k$.

The repeated initialization of the covariance matrix at each step is necessary to prevent its unbounded growth, as we assume that the system evolves without feedback control.
In the absence of feedback, the uncertainty on the position and orientation of the body frame accumulates along the track, which may severely and unrealistically degrade the conditioning of the covariance matrix. 
For this reason, the propagation horizon in this open-loop setting is a critical design parameter that must be bounded to realistically capture the evolution of disturbances occurring prior to any corrective driver response.

This approach requires modifications to the formulation~\eqref{eq:DOCP} introduced in Section~\ref{sec:discretization}. In particular, Eqs.~\eqref{eq:DOCPdP}, \eqref{eq:DOCPdPIC}, and \eqref{eq:DOCPconstraints} are replaced by Eqs.~\eqref{eq:DOCPdPopenloop}, \eqref{eq:DOCPdPinitopenloop}, and \eqref{eq:DOCPconstraintsopenloop}. Equation~\eqref{eq:DOCPdPopenloop} collects the continuity and collocation equations for all versions of the covariance matrix propagated from previous steps. To support the evolution of $H$ versions, we introduce $H \times d$ matrices $\bSi_k^j = (\bSi_{k,1}^j, \ldots, \bSi_{k,d}^j)$, representing the collocation values of each version $j$ of the covariance matrix within interval $k$.

%The proposed approach requires the presence of $H+1$ versions of the matrix $\bP$ at each grid point of the discretized problem. One of these versions is initialized with $\bar{\bP}_0$, while the others represent the evolution of matrices that were initialized at previous grid points, from 1 to $H$ steps earlier. This ensures that at each step it is possible to find a covariance matrix evolved for $H$ steps and use it to robustify the constraints. The initialization of the covariance matrix is necessary due to its divergent dynamics. In fact, without a feedback control, the error on the position and orientation of the body reference frame can only increase along the track, compromising the conditioning of the matrix.
%
%Introducing the versioning of the covariance matrix approach requires a modification of the Equations described in Section~\ref{sec:discretization}. In particular Eqs.~\eqref{eq:DOCPdP}, \eqref{eq:DOCPdPIC}, and \eqref{eq:DOCPconstraints} are replaced by Eqs.~\eqref{eq:DOCPdPopenloop}, \eqref{eq:DOCPdPinitopenloop}, and \eqref{eq:DOCPconstraintsopenloop}. Eq.~\eqref{eq:DOCPdPopenloop} collects the continuity and collocation Equations for all the versions of the covariance matrix that have propagated from the previous step. To account for the evolution of $H$ matrices, it is necessary to introduce $H\times d$ matrices $\bSi_k^j=(\bSi_{k,1}^j, \ldots, \bSi_{k,d}^j)$, which represent the values of each matrix version at the collocation points.

The specific formulation is expressed as follows:
\begin{subequations}\label{eq:DOCPopenloop}
\begin{alignat}{3}
	\underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k}{\text{minimize}} \,
	& & & J_k(\bmu_k,\bxi_k, \bu_k) & & \label{eq:DOCPcostopenloop} \\
	\hspace*{-2.0 cm}\text{s.t.} \quad
	& \bzero      & = & \; \bPsimu_k(\bmu_{k-1},\bmu_k,\bxi_k, \bu_k,\bz_k),
	& \quad & k = 1,\ldots, N \label{eq:DOCPdynopenloop} \\
	& \bmu_0      & = & \; \bar{\bmu}_0
	& & \label{eq:DOCPdynICopenloop} \\
	& \bzero      & = & \; \bPsiP_k(\bmu_k,\bxi_k, \bu_k, \bP_{k-1}^{j-1},\bP_k^j,\bSi_k^j,\bz_k),
	& \quad & \substack{k = 1,\ldots, N; \; \\ j=1,\ldots,H;\;} \label{eq:DOCPdPopenloop} \\
	& \bP_k^0       & = & \; \bar{\bP}_0 \succeq 0
	& \quad & k = 1,\ldots, N; \; \label{eq:DOCPdPinitopenloop} \\
	& \bzero      & = & \; \bOm_k(\bmu_k,\bxi_k, \bu_k,\bz_k),
	& \quad & k = 0,\ldots, N \label{eq:DOCPpathopenloop} \\
	& 0           & \geq & \; h_i(\bmu_k, \bu_k, \bz_k) + \be_i(\bmu_k, \bu_k, \bP_k^H, \bz_k),
	& \quad & k = 1,\ldots, N;\; i \in \calI \label{eq:DOCPconstraintsopenloop}
\end{alignat}
\end{subequations}
Equation~\eqref{eq:DOCPdPinitopenloop} defines the initialization of the appropriate covariance matrix version at step $k$, while Eq.~\eqref{eq:DOCPconstraintsopenloop} specifies that the back-off term in the constraints is computed using the most propagated version, $\bP^H_k$. A schematic illustration of the management of the $H+1$ covariance matrix instances is provided in Figure~\ref{fig:DOCPgrid}. The dashed rectangle highlights the $k$-th discretization step, where the constraints are evaluated. At each grid point, two particular versions of the covariance matrix are emphasized: $\bP^0_k$ (red node), representing the instance to be initialized at that step, and $\bP^H_k$ (green node), corresponding to the version that has been propagated over the full horizon of $H$ steps and is used to robustify the constraints in Eq.~\eqref{eq:DOCPconstraintsopenloop}.

%Eq.~\eqref{eq:DOCPdPinitopenloop} is used to initialize the correct version at the $k$-th step while Eq.~\eqref{eq:DOCPconstraintsopenloop} specify that the back-off term of the constraints is evaluated with the most propagated covariance matrix version, $\bP^H_k$.
%A schematic representation of the approach used to manage the $H+1$ versions of the covariance matrix is shown in Figure \ref{fig:DOCPgrid}, where the dashed rectangle represent the $k$-th step at which the constraints are being formulated. At each grid point, two versions of the covariance matrix are highlighted: $\bP^0_k$ (red node), which is the version to be initialized, and $\bP^H_k$ (green node), which is the most propagated version used to formulate the robust constraints in Eq.~\eqref{eq:DOCPconstraintsopenloop}.

@@

	\end{tikzpicture}
\caption{Schematic representation of the continuity and initialization equations associated with the $H+1$ propagated instances of the covariance matrix introduced in problem~\eqref{eq:DOCPopenloop}. The dashed rectangle highlights the $k$-th discretization step, at which the robust constraints are enforced. Each node corresponds to a specific covariance matrix, where the subscript denotes the current grid index and the superscript indicates the number of propagation steps since initialization. At each step, two key instances are emphasized: the matrix to be initialized at that step (red node), and the matrix that has been propagated over $H$ steps (green node), which is used in the evaluation of the constraint backoff.}
	%\caption{Schematic representation of the continuity and initialization Equations for the $H+1$ versions of the covariance matrix introduced in the discretized OCP. The dashed rectangle indicate the $k$-th step, in which the constraints are formulated. Each node represents a covariance matrix, labeled such that the subscript denotes the current grid step, while the superscript indicates the number of propagation steps it has undergone. At each step, two nodes are highlighted: the covariance matrix to be initialized (red node) and the matrix that has been propagated for $H$ steps (green node), which is used to formulate the robust constraints.}
	\label{fig:DOCPgrid}
\end{figure}
